---
title: "STA5092Z Assignment 1"
author: "Tino Muzambi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r, echo=FALSE}
#install.packages("tidyverse")
library(tidyverse)
```

# Read in data.

```{r}
# A1 & A2
a12.2015 <- read.csv("./Data/A1_2_2015.csv") %>% as_tibble
a12.2016 <- read.csv("./Data/A1_2_2016.csv") %>% as_tibble
a12.2017 <- read.csv("./Data/A1_2_2017.csv") %>% as_tibble
a12.2018 <- read.csv("./Data/A1_2_2018.csv") %>% as_tibble
a12.2019 <- read.csv("./Data/A1_2_2019.csv") %>% as_tibble
a12.2020 <- read.csv("./Data/A1_2_2020.csv") %>% as_tibble
a12.2021 <- read.csv("./Data/A1_2_2021.csv") %>% as_tibble
a12.2022 <- read.csv("./Data/A1_2_2022.csv") %>% as_tibble

# B1
b1.2015 <- read.csv("./Data/B1_2015.csv") %>% as_tibble
b1.2016 <- read.csv("./Data/B1_2016.csv") %>% as_tibble
b1.2017 <- read.csv("./Data/B1_2017.csv") %>% as_tibble
b1.2018 <- read.csv("./Data/B1_2018.csv") %>% as_tibble
b1.2019 <- read.csv("./Data/B1_2019.csv") %>% as_tibble
b1.2020 <- read.csv("./Data/B1_2020.csv") %>% as_tibble
b1.2021 <- read.csv("./Data/B1_2021.csv") %>% as_tibble
b1.2022 <- read.csv("./Data/B1_2022.csv") %>% as_tibble

# B2
b2.2015 <- read.csv("./Data/B2_2015.csv") %>% as_tibble
b2.2016 <- read.csv("./Data/B2_2016.csv") %>% as_tibble
b2.2017 <- read.csv("./Data/B2_2017.csv") %>% as_tibble
b2.2018 <- read.csv("./Data/B2_2018.csv") %>% as_tibble
b2.2019 <- read.csv("./Data/B2_2019.csv") %>% as_tibble
b2.2020 <- read.csv("./Data/B2_2020.csv") %>% as_tibble
b2.2021 <- read.csv("./Data/B2_2021.csv") %>% as_tibble
b2.2022 <- read.csv("./Data/B2_2022.csv") %>% as_tibble
```

# Checking the data.

```{r, echo=FALSE}
print("A1_2")
str(a12.2015)
str(a12.2016)
str(a12.2017)
str(a12.2018)
str(a12.2019)
str(a12.2020)
str(a12.2021)
str(a12.2022)

print("B1")
str(b1.2015)
str(b1.2016)
str(b1.2017)
str(b1.2018)
str(b1.2019)
str(b1.2020)
str(b1.2021)
str(b1.2022)

print("B2")
str(b2.2015)
str(b2.2016)
str(b2.2017)
str(b2.2018)
str(b2.2019)
str(b2.2020)
str(b2.2021)
str(b2.2022)
```
# Merge into one tibble.

```{r}
# First rename StudentID variable to ID
a12.2015 <- a12.2015 %>%
  rename(ID = StudentID)
```

```{r}
undergraduate <- bind_rows(list(a12.2015, a12.2016, a12.2017, a12.2018, a12.2019, a12.2020, a12.2021, a12.2022,
                                b1.2015, b1.2016, b1.2017, b1.2018, b1.2019, b1.2020, b1.2021, b1.2022,
                                b2.2015, b2.2016, b2.2017, b2.2018, b2.2019, b2.2020, b2.2021, b2.2022))
str(undergraduate)
```

# Rename variables to more descriptive names.

```{r}
undergraduate <- undergraduate %>% 
  rename(year = Term,
         course = Catalog.Nbr,
         degree = Acad.Prog,
         id = ID,
         grade = Grade)
str(undergraduate)
```



# Look at top and bottom of data.

```{r}
head(undergraduate)
tail(undergraduate)
```

# Check total number of observations.

```{r}
dim(undergraduate)
```

# Create new grade column which has integers only.

```{r}
# If string value in grade is an integer, convert to integer, else check if it's UP, if so, replace with 50, else replace with 0.
undergraduate <- undergraduate %>%
            mutate(num.grade = ifelse(grepl("^[0-9]+$", grade), as.integer(grade), ifelse(grade == "UP" | grade == "PA", 50, 0)))
str(undergraduate)
```

# 1. What is the total number and percentage of students for each year who pass both the 1st and the 2nd year courses in the same year?

```{r}
# Group data by year and find unique students who pass both the 1st and the 2nd year courses in the same year.
pass.both <- undergraduate %>%
  group_by(year) %>%
  mutate(total.number = n()) %>%
  filter(num.grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.both

# Plot the results.
ggplot(pass.both, aes(x = as.factor(year), y = perc.pass)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Students Passing Both 1st and 2nd Year Courses",
       x = "Year",
       y = "Percentage of Students Passing (%)") +
  theme(plot.title = element_text(hjust = 0.5))
```

# 2. Are there any outliers in the grade variable? What is your definition of an outlier?

```{r}
# Plot box and whisker plot to view grade distribution.
grade.box <- undergraduate %>%
  ggplot(aes(num.grade)) +
  geom_boxplot()
grade.box

# Plot histogram to view grade distribution.
grade.hist <- undergraduate %>%
  ggplot(aes(num.grade)) +
  geom_histogram()
grade.hist

# Use summary to view grade distribution.
undergraduate %>%
  summary()
```

# 3. Are there any obvious clusters in the grade variable? i.e. more marks around 50%.

```{r}
# Plot density plot to view clustering of grades.
grade.density <- undergraduate %>% 
  ggplot(aes(num.grade)) +
  geom_density()
grade.density
```

# 4. What is the pass rate for the 1st and 2nd year courses for each year? Are there any differences between A1 and A2?

```{r}
# Calculate pass rate for 1st and 2nd year courses group by year.
pass.rates.first.second <- undergraduate %>%
  mutate(first.second = ifelse(course == "A1" | course == "A2", "First", "Second")) %>% 
  group_by(first.second, year) %>% 
  mutate(total.number = n()) %>%
  filter(num.grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.rates.first.second

# Plot the results.
ggplot(pass.rates.first.second, aes(x = as.factor(year), y = perc.pass, fill = first.second)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Pass Rates for 1st and 2nd Year Courses Grouped by Year",
       x = "Year",
       y = "Percentage of Students Passing (%)",
       fill = "Year") +
  scale_fill_manual(values = c("First" = "blue", "Second" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Calculate pass rate for A1 and A2 group by year.
pass.rates.a12 <- undergraduate %>%
  group_by(course, year) %>% 
  filter(course == "A1" | course == "A2") %>% 
  mutate(total.number = n()) %>%
  filter(num.grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.rates.a12

# Plot the results.
ggplot(pass.rates.a12, aes(x = as.factor(year), y = perc.pass, fill = course)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Pass Rates for A1 and A2 Courses Grouped by Year",
       x = "Year",
       y = "Percentage of Students Passing (%)",
       fill = "Course") +
  scale_fill_manual(values = c("A1" = "blue", "A2" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# 5. How many students fail B1 even if they pass A1 or A2?

```{r}
# Get unique list of students who passed A1 or A2.
pass.a12 <- undergraduate %>% 
  filter(course %in% c("A1", "A2") & num.grade >= 50) %>% 
  distinct(id, .keep_all=TRUE)

# Get unique list of students who failed B1.
fail.b1 <- undergraduate %>% 
  filter(course == "B1" & num.grade < 50) %>% 
  distinct(id)

# Find intersection to get students who failed B1 even if they passed A1 or A2.
pass.a12.fail.b1 <- intersect(pass.a12$id, fail.b1$id) %>% 
  length()
pass.a12.fail.b1

# Get number of students who took B1.
num.b1 <- undergraduate %>% 
  filter(course == "B1") %>% 
  distinct(id) %>% 
  nrow()

# Calculate percentage of students who failed B1 even if they passed A1 or A2.
b1.perc.fail <- (pass.a12.fail.b1 / num.b1) * 100
b1.perc.fail
```

# 6. What would you suggest the minimum mark obtained should be from 1st year courses so that the student is able to pass B1?

```{r}
# Get unique list of students who passed B1.
pass.b1 <- undergraduate %>% 
  filter(course == "B1" & num.grade >= 50) %>%
  distinct(id, .keep_all=TRUE)

# Get list of students who passed A1 or A2 and B1.
pass.a12.pass.b1 <- right_join(pass.a12, pass.b1, by = "id")

# Find minimum mark
min.mark <- min(pass.a12.pass.b1$num.grade.x, na.rm = TRUE)
min.mark
```

# 7. What would you suggest the minimum mark obtained should be for B1 so that the student is able to pass B2?

```{r}
# Get unique list of students who passed B2.
pass.b2 <- undergraduate %>% 
  filter(course == "B2" & num.grade >= 50) %>%
  distinct(id)

# Get intersection of those who passed B1 and B2.
pass.b1.b2 <- intersect(pass.b1$id, pass.b2$id)

# Get students who are in this intersection.
pass.b1.b2.data <- undergraduate %>%
  filter(id %in% pass.b1.b2)

# Find minimum mark.
min_mark.b1.b2 <- pass.b1.b2.data %>%
  filter(course == "B1") %>%
  summarise(min.mark = min(num.grade, na.rm = TRUE))
min_mark.b1.b2
```


# 8. Is there any hope for students with a UP (supplementary exam) for A1 or A2 to make it to the 3rd year of their studies without failing B1 and/or B2?

```{r}
# Identify students with a UP for A1 or A2.
up.a1.a2 <- undergraduate %>%
  filter((course %in% c("A1", "A2")) & (grade == "UP")) %>%
  distinct(id)

# Check if these students pass B1 and B2.
pass.b1.b2.up.a1.a2 <- undergraduate %>%
  filter(id %in% up.a1.a2$id & (course %in% c("B1", "B2")) & num.grade >= 50) %>%
  distinct(id)

# Calculate the number of students who passed B1 and B2 after having a UP for A1 or A2.
num.students <- nrow(pass.b1.b2.up.a1.a2)

# Calculate the percentage of students who passed B1 and B2 after having a UP for A1 or A2.
perc.pass <- (num.students / nrow(up.a1.a2)) * 100
perc.pass
```

# 9. What is the correlation between the grades of 1st and 2nd year courses?

```{r}
# Pivot the data to have 1st year and 2nd year grades as columns.
grades.pivot <- undergraduate %>%
  pivot_wider(names_from = course, values_from = num.grade)

# Filter out rows with missing values in A1 or B1.
filtered.data <- grades.pivot %>%
  filter(!is.na(A1) | !is.na(B1))

# Calculate the correlation between 1st and 2nd-year grades.
cor(filtered.data$A1, filtered.data$B1)
```

# 10. Determine if answers for any of the above questions have been affected by Covid years? UCT has experienced other disasters, protests. Do you see any impact of these on your answers from one year to another?

```{r}
# Compare pass rates between years.
pass.rates.comparison <- undergraduate %>%
  group_by(course, year) %>%
  summarise(mean.pass.rate = mean(num.grade >= 50) * 100)
pass.rates.comparison

# Plot the comparison.
pass.rates.comparison %>% 
  ggplot(aes(x = as.factor(year), y = mean.pass.rate, fill = as.factor(year == 2020))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"), guide = "none") +
  labs(title = "Pass Rates Comparison - Covid Year vs Non-Covid Years",
       x = "Year",
       y = "Average Pass Rate (%)") +
  theme_minimal()
```

# 11. What kind of an informed suggestion/advice one can make to a student who is registering for A1 or A2 for the first time? What would you look at to decide from the very beginning that the student will be successful in these courses?

```{r}

```

# 12. Are there any patterns or trends in the academic progress of students across different degrees over the years?

```{r}
# Group data by year and degree, calculate average grade and pass rates.
degree.analysis <- undergraduate %>%
  group_by(year, degree) %>%
  summarise(mean.grade = mean(num.grade, na.rm = TRUE),
            pass.rate = mean(num.grade >= 50, na.rm = TRUE) * 100)

# Plot the average grade.
p.mean.grade <- degree.analysis %>% 
  ggplot(aes(x = as.factor(year), y = mean.grade, color = degree)) +
  geom_point() +
  geom_line(aes(group = degree), linetype = "dashed") +
  facet_wrap(~degree, scales = "free_y") +
  labs(title = "Average Grade for Different Degrees Grouped by Year",
       x = "Year",
       y = "Average Grade",
       color = "Degree") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

# Plot the pass rate.
p.pass.rate <- degree.analysis %>% 
  ggplot(aes(x = as.factor(year), y = pass.rate, fill = degree)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~degree, scales = "free_y") +
  labs(title = "Pass Rates for Different Degrees Grouped by Year",
       x = "Year",
       y = "Pass Rate (%)",
       fill = "Degree") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

degree.analysis
```

# 13. Can we identify any patterns or trends in the success rate of students who faced challenges in their first year but later excelled in subsequent years?

```{r}
# Identify students who faced challenges in their first year.
challenged.first.year <- undergraduate %>%
  filter(year %in% c(2015, 2016, 2017) & grade %in% c("DPR", "INC", "AB"))

# Identify students who excelled in subsequent years.
excelled.subsequent.years <- undergraduate %>%
  filter(year %in% c(2018, 2019, 2020, 2021, 2022) & num.grade >= 50)

# Find students who faced challenges in the first year but later excelled.
resilient.students <- inner_join(challenged.first.year, excelled.subsequent.years, by = "id")
resilient.students
```

