---
title: "STA5092Z Assignment 1"
author: "Tino Muzambi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    highlight: pygments
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r, echo=FALSE}
#install.packages("tidyverse")
library(tidyverse)
library(naniar)
```

# Read in data.

```{r}
# A1 & A2
a12.2015 <- read.csv("./Data/A1_2_2015.csv") %>% as_tibble
a12.2016 <- read.csv("./Data/A1_2_2016.csv") %>% as_tibble
a12.2017 <- read.csv("./Data/A1_2_2017.csv") %>% as_tibble
a12.2018 <- read.csv("./Data/A1_2_2018.csv") %>% as_tibble
a12.2019 <- read.csv("./Data/A1_2_2019.csv") %>% as_tibble
a12.2020 <- read.csv("./Data/A1_2_2020.csv") %>% as_tibble
a12.2021 <- read.csv("./Data/A1_2_2021.csv") %>% as_tibble
a12.2022 <- read.csv("./Data/A1_2_2022.csv") %>% as_tibble

# B1
b1.2015 <- read.csv("./Data/B1_2015.csv") %>% as_tibble
b1.2016 <- read.csv("./Data/B1_2016.csv") %>% as_tibble
b1.2017 <- read.csv("./Data/B1_2017.csv") %>% as_tibble
b1.2018 <- read.csv("./Data/B1_2018.csv") %>% as_tibble
b1.2019 <- read.csv("./Data/B1_2019.csv") %>% as_tibble
b1.2020 <- read.csv("./Data/B1_2020.csv") %>% as_tibble
b1.2021 <- read.csv("./Data/B1_2021.csv") %>% as_tibble
b1.2022 <- read.csv("./Data/B1_2022.csv") %>% as_tibble

# B2
b2.2015 <- read.csv("./Data/B2_2015.csv") %>% as_tibble
b2.2016 <- read.csv("./Data/B2_2016.csv") %>% as_tibble
b2.2017 <- read.csv("./Data/B2_2017.csv") %>% as_tibble
b2.2018 <- read.csv("./Data/B2_2018.csv") %>% as_tibble
b2.2019 <- read.csv("./Data/B2_2019.csv") %>% as_tibble
b2.2020 <- read.csv("./Data/B2_2020.csv") %>% as_tibble
b2.2021 <- read.csv("./Data/B2_2021.csv") %>% as_tibble
b2.2022 <- read.csv("./Data/B2_2022.csv") %>% as_tibble
```

# Checking the data.

```{r, echo=FALSE}
print("A1_2")
str(a12.2015)
str(a12.2016)
str(a12.2017)
str(a12.2018)
str(a12.2019)
str(a12.2020)
str(a12.2021)
str(a12.2022)

print("B1")
str(b1.2015)
str(b1.2016)
str(b1.2017)
str(b1.2018)
str(b1.2019)
str(b1.2020)
str(b1.2021)
str(b1.2022)

print("B2")
str(b2.2015)
str(b2.2016)
str(b2.2017)
str(b2.2018)
str(b2.2019)
str(b2.2020)
str(b2.2021)
str(b2.2022)
```
# Merge into one tibble.

```{r}
# First rename StudentID variable to ID
a12.2015 <- a12.2015 %>%
  rename(ID = StudentID)
```

```{r}
undergraduate <- bind_rows(list(a12.2015, a12.2016, a12.2017, a12.2018, a12.2019, a12.2020, a12.2021, a12.2022,
                                b1.2015, b1.2016, b1.2017, b1.2018, b1.2019, b1.2020, b1.2021, b1.2022,
                                b2.2015, b2.2016, b2.2017, b2.2018, b2.2019, b2.2020, b2.2021, b2.2022))
str(undergraduate)
```

# Look at top and bottom of data.

```{r}
head(undergraduate)
tail(undergraduate)
```

# Check total number of observations.

```{r}
dim(undergraduate)
```

# Rename variables to more descriptive names.

```{r}
undergraduate <- undergraduate %>% 
  rename(year = Term,
         course = Catalog.Nbr,
         degree = Acad.Prog,
         id = ID,
         grade = Grade)
str(undergraduate)
```

# Create new code column which has pass codes only and only keep numeric grades in grade column.

```{r}
# Create a new column with codes and NAs for numerical grades.
undergraduate <- undergraduate %>%
  mutate(code = ifelse(grepl("^[0-9]+$", grade), NA, grade))

# Create vectors of passing and failing codes.
failing.codes <- c("DPR", "DE","INC", "AB", "F", "LOA")
passing.codes <- c("UP","PA")

# Replace codes with 0s and 50s
undergraduate <- undergraduate %>%
  mutate(grade = ifelse(grade %in% failing.codes, 0,
                        ifelse(grade %in% passing.codes, 50, 
                               ifelse(grepl("^[0-9]+A$", grade),0,grade)))
  )

# Convert grade to numerical type.
undergraduate <- undergraduate %>%
  mutate(grade = as.numeric(grade, errors = 'coerce'))

str(undergraduate)
```

# 1. What is the total number and percentage of students for each year who pass both the 1st and the 2nd year courses in the same year?

```{r}
# Group data by year and find unique students who pass both the 1st and the 2nd year courses in the same year.
pass.both <- undergraduate %>%
  group_by(year) %>%
  mutate(total.number = n()) %>%
  filter(grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.both

# Plot the results.
ggplot(pass.both, aes(x = as.factor(year), y = perc.pass)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Students Passing Both 1st and 2nd Year Courses",
       x = "Year",
       y = "Percentage of Students Passing (%)") +
  theme(plot.title = element_text(hjust = 0.5))
```

# 2. Are there any outliers in the grade variable? What is your definition of an outlier?

```{r}
# Use summary to view grade distribution.
undergraduate$grade %>%
  summary()

# Plot box and whisker plot to view grade distribution.
grade.box <- undergraduate %>%
  ggplot(aes(grade)) +
  geom_boxplot() +
  labs(title = "Boxplot of Grade",
       x = "Grade (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
grade.box
```

# 3. Are there any obvious clusters in the grade variable? i.e. more marks around 50%.

```{r}
# Plot density plot to view clustering of grades.
grade.density <- undergraduate %>% 
  ggplot(aes(grade)) +
  geom_density() +
  labs(title = "Density of Grade",
       x = "Grade (%)",
       y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
grade.density

# Plot histogram to view grade distribution.
grade.hist <- undergraduate %>%
  ggplot(aes(grade)) +
  geom_histogram() +
  labs(title = "Histogram of Grade Distribution",
       x = "Grade (%)",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
grade.hist
```

# 4. What is the pass rate for the 1st and 2nd year courses for each year? Are there any differences between A1 and A2?

```{r}
# Calculate pass rate for 1st and 2nd year courses group by year.
pass.rates.first.second <- undergraduate %>%
  mutate(first.second = ifelse(course == "A1" | course == "A2", "First", "Second")) %>% 
  group_by(first.second, year) %>% 
  mutate(total.number = n()) %>%
  filter(grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.rates.first.second

# Plot the results.
plot.pass.rates.first.second <- ggplot(pass.rates.first.second, aes(x = as.factor(year), y = perc.pass, fill = first.second)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(perc.pass, 1), vjust = 1.5, hjust = 0.5), position = position_dodge(width = 0.9), color = "white") +
  labs(title = "Pass Rates for 1st and 2nd Year Courses Grouped by Year",
       x = "Year",
       y = "Percentage of Students Passing (%)",
       fill = "Year") +
  scale_fill_manual(values = c("First" = "blue", "Second" = "red"), name = "Course Year") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
plot.pass.rates.first.second

# Calculate pass rate for A1 and A2 group by year.
pass.rates.a12 <- undergraduate %>%
  group_by(course, year) %>% 
  filter(course == "A1" | course == "A2") %>% 
  mutate(total.number = n()) %>%
  filter(grade >= 50) %>%
  distinct(id, .keep_all = TRUE) %>%
  summarise(total.pass = n(), perc.pass = (n() / first(total.number)) * 100)
pass.rates.a12

# Plot the results.
plot.pass.rates.a12 <- ggplot(pass.rates.a12, aes(x = as.factor(year), y = perc.pass, fill = course)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(perc.pass, 1), vjust = 1.5, hjust = 0.5), position = position_dodge(width = 0.9), color = "white") +
  labs(title = "Pass Rates for A1 and A2 Courses Grouped by Year",
       x = "Year",
       y = "Percentage of Students Passing (%)",
       fill = "Course") +
  scale_fill_manual(values = c("A1" = "blue", "A2" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
plot.pass.rates.a12
```

# 5. How many students fail B1 even if they pass A1 or A2?

```{r}
# Get unique list of students who passed A1 or A2.
pass.a12 <- undergraduate %>% 
  filter(course %in% c("A1", "A2") & grade >= 50) %>% 
  distinct(id, .keep_all=TRUE)

# Get unique list of students who failed B1.
fail.b1 <- undergraduate %>% 
  filter(course == "B1" & grade < 50) %>% 
  distinct(id)

# Find intersection to get students who failed B1 even if they passed A1 or A2.
pass.a12.fail.b1 <- intersect(pass.a12$id, fail.b1$id) %>% 
  length()
cat("Number of students who failed B1 even if they passed A1 or A2:", pass.a12.fail.b1, "\n")


# Get number of students who took B1.
num.b1 <- undergraduate %>% 
  filter(course == "B1") %>% 
  distinct(id) %>% 
  nrow()

# Calculate percentage of students who failed B1 even if they passed A1 or A2.
b1.perc.fail <- (pass.a12.fail.b1 / num.b1) * 100
cat("Percentage of students who failed B1 even if they passed A1 or A2:", round(b1.perc.fail, 2), "%\n")
```


# 6. What would you suggest the minimum mark obtained should be from 1st year courses so that the student is able to pass B1?

```{r}
# Filter to only 1st year and B1 courses.
a1.a2.b1 <- undergraduate %>% 
  filter(course %in% c("A1", "A2","B1"))

# Join data to link 1st and 2nd year grades by student.
a1.a2.b1.joined <- merge(subset(a1.a2.b1, course %in% c("A1", "A2")),  
                   subset(a1.a2.b1, course=="B1"), 
                   by="id")
                   
# Take max of A1 and A2 grades for each student.
a1.a2.b1.joined$a.grade <- pmax(a1.a2.b1.joined$grade.x, a1.a2.b1.joined$grade.y)


# Identify the minimum 1st year grade passed by 75% of students who passed B1.
print("Minimum suggested mark from 1st year so that the student is able to pass B1:")
quantile(a1.a2.b1.joined$a.grade[a1.a2.b1.joined$grade.y >= 50], 0.25, na.rm = TRUE)
```

# 7. What would you suggest the minimum mark obtained should be for B1 so that the student is able to pass B2?

```{r}
# Filter to only B1 and B2 courses.
b12 <- undergraduate %>% 
  filter(course %in% c("B1", "B2"))

# Join data to link B1 and B2 grades by student.
b1.b2.joined <- merge(subset(b12, course=="B1"), 
                   subset(b12, course=="B2"),
                   by="id")

# Identify the minimum B1 grade passed by 75% of students who passed B2.
print("Minimum suggested mark from B1 so that the student is able to pass B2:")
quantile(b1.b2.joined$grade.x[b1.b2.joined$grade.y >= 50], 0.25, na.rm = TRUE)
```


# 8. Is there any hope for students with a UP (supplementary exam) for A1 or A2 to make it to the 3rd year of their studies without failing B1 and/or B2?

```{r}
# Identify students with a UP for A1 or A2.
up.a1.a2 <- undergraduate %>%
  filter((course %in% c("A1", "A2")) & (code == "UP")) %>%
  distinct(id)

# Check if these students pass B1 and B2.
pass.b1.b2.up.a1.a2 <- undergraduate %>%
  filter(id %in% up.a1.a2$id & (course %in% c("B1", "B2")) & grade >= 50) %>%
  distinct(id)

# Calculate the number of students who passed B1 and B2 after having a UP for A1 or A2.
num.students <- nrow(pass.b1.b2.up.a1.a2)

# Calculate the percentage of students who passed B1 and B2 after having a UP for A1 or A2.
perc.pass.b12.up.a12 <- (num.students / nrow(up.a1.a2)) * 100
cat("Percentage of students who passed B1 and B2 after having a UP for A1 or A2:", round(perc.pass.b12.up.a12, 2), "%")
```

# 9. What is the correlation between the grades of 1st and 2nd year courses?

```{r}
# Separate the grades for 1st and 2nd year courses.
a12 <- undergraduate %>% 
  filter(course %in% c("A1", "A2"))
b12 <- undergraduate %>% 
  filter(course %in% c("B1", "B2"))

# Join data to link 1st and 2nd year grades by student.
merged.joined <- merge(a12, 
                       b12, 
                       by = "id", 
                       suffixes = c(".first", ".second"))

# Calculate the correlation.
correlation.first.second <- cor(merged.joined$grade.first, merged.joined$grade.second, use = "pairwise.complete.obs")
cat("Correlation between the grades of 1st and 2nd year courses:", round(correlation.first.second, 2))
```


# 10. Determine if answers for any of the above questions have been affected by Covid years? UCT has experienced other disasters, protests. Do you see any impact of these on your answers from one year to another?

```{r}
# Compare pass rates between years.
pass.rates.comparison <- undergraduate %>%
  group_by(year) %>%
  summarise(mean.pass.rate = mean(grade >= 50) * 100)
pass.rates.comparison

# Plot the comparison.
plot.pass.rates.comparison <- pass.rates.comparison %>% 
  ggplot(aes(x = as.factor(year), y = mean.pass.rate, fill = as.factor(year == 2020))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(mean.pass.rate, 1), vjust = 1.5, hjust = 0.5), position = position_dodge(width = 0.9), color = "white") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "grey"), guide = "none") +
  labs(title = "Pass Rates Comparison - Covid Year vs Non-Covid Years",
       x = "Year",
       y = "Average Pass Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
plot.pass.rates.comparison
```

# 11. What kind of an informed suggestion/advice one can make to a student who is registering for A1 or A2 for the first time? What would you look at to decide from the very beginning that the student will be successful in these courses?

```{r}

```

# 12. Are there any patterns or trends in the academic progress of students across different degrees over the years?

```{r}
# Group data by year and degree, calculate average grade and pass rates.
degree.analysis <- undergraduate %>%
  group_by(year, degree) %>%
  summarise(mean.grade = mean(grade, na.rm = TRUE),
            pass.rate = mean(grade >= 50, na.rm = TRUE) * 100)
degree.analysis

# Plot the average grade.
p.mean.grade <- degree.analysis %>% 
  ggplot(aes(x = as.factor(year), y = mean.grade)) +
  geom_point() +
  geom_line(aes(group = degree)) +
  scale_y_continuous(limits = c(0, 100)) + 
  facet_wrap(~degree, scales = "free_y") +
  labs(title = "Average Grade for Different Degrees Grouped by Year",
       x = "Year",
       y = "Average Grade (%)",
       color = "Degree") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
p.mean.grade

# Plot the pass rate.
p.pass.rate <- degree.analysis %>% 
  ggplot(aes(x = as.factor(year), y = pass.rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(limits = c(0, 100)) + 
  facet_wrap(~degree, scales = "free_y") +
  labs(title = "Pass Rates for Different Degrees Grouped by Year",
       x = "Year",
       y = "Pass Rate (%)",
       fill = "Degree") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
p.pass.rate
```

# 13. Can we identify any patterns or trends in the success rate of students who faced challenges in their first year but later excelled in subsequent years?

```{r}
# Identify students who faced challenges in earlier years.
challenged.first.year <- undergraduate %>%
  filter(year %in% c(2015, 2016, 2017) & code %in% failing.codes) %>% 
  select(year, course, grade, id, code)

# Identify students who excelled in subsequent years.
excelled.subsequent.years <- undergraduate %>%
  filter(year %in% c(2018, 2019, 2020, 2021, 2022) & grade >= 50) %>% 
  select(year, course, grade, id, code)

# Find students who faced challenges in the first year but later excelled.
resilient.students <- inner_join(challenged.first.year, excelled.subsequent.years, by = "id", suffix = c(".challenged", ".excelled"))
resilient.students
```

